Direction-aware functional class scoring enrichment analysis of Infinium DNA methylation data

### 2024-01-24 manuscript/manuscript.Rmd
update

<strong>
Mark Ziemann<sup>1,2*</sup>,
Mandhri Abeysooriya<sup>2,3</sup>,
Anusuiya Bora<sup>1,2</sup>,
Severine Lamon<sup>3</sup>,
Mary Sravya Kasu<sup>1</sup>,
Mitch W. Norris<sup>1</sup>,
Yen Ting Wong<sup>4</sup>,
Jeffrey Craig<sup>4,5</sup>
</strong>

@@

Infinium methylation beadarrays remain one of the most popular platforms for studying genomic
DNA methylation, but tools for downstream pathways analysis have their limitations.
Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of
genes and evaluation of their collective regulation in biological systems, but the implementations described for
Infinium methylation data do not retain direction information which is important for mechanistic understanding
of genomic regulation.
Here, we evaluate several candidate FCS methods that retain directional information.
According to simulation results, the best performing method involves the mean aggregation of probe limma
t-statistics by gene followed by a rank-ANOVA enrichment test using the mitch package.
This method, which we call "LAM", outperformed an existing over-representation analysis method in
simulations, and showed higher sensitivity and robustness in an analysis of real lung tumor-normal paired datasets.
Using matched RNA-seq data we examine the relationship of methylation differences at promoters and
gene bodies with RNA expression at the level of pathways in lung cancer.
To further demonstrate the utility of our approach, we apply it to two other contexts; firstly an examination of
methylation differences associated with chronological age; and secondly to examine methylation differences in
infants conceived with in vitro fertilisation.
These results indicate that the LAM method is more sensitive and precise as compared to existing methods.
LAM has the potential to illuminate disease mechanisms and identify novel diagnostic signatures.
A vignette is provided to illustrate how to implement this method.

@@

DNA methylation is the most widely studied epigenetic mark, for its role in
development and disease [@Robertson2005-tu].
Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
methylation patterns in disease and other contexts [@Li2019-og].
Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
as compared to high throughput methylation sequencing [@Wei2021-qj].
Infinium arrays typically have multiple probes per gene in different locations including promoters,
CpG islands, gene bodies and enhancers [@Pidsley2016-ix], which complicates downstream functional enrichment analysis,
otherwise known as pathway analysis or gene set enrichment analysis.
Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
class scoring (FCS) [@Khatri2012-qy].
A third type of enrichment analysis called pathway topology improves upon other methods with more sophisticated
modelling of gene network activations, but these are yet to be adopted widely [@Xie2021-py].
In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
to a background list consisting of genes measured in the assay.
The test seeks to identify sets of genes (eg: ontologies, pathways) that are over-represented in the
gene list of interest relative to the background [@Tavazoie1999-ng].
FCS takes a different approach by ranking all detected genes by a differential
regulation score (eg: fold change, confident effect size) followed by a test to assess whether each
set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of infinium methylation array
data as it addresses issues related to probes belonging to more than one gene and the fact that genes can have multiple
probes [@Maksimovic2021-ur].
As gsameth is an ORA method, results dependent strongly on the significance threshold used [@Ren2019-mr], and also
on the proportion of probes that meet this threshold.
FCS methods `ebGSEA` and `methylGSA` have been developed for EWAS data and are suggested to have better sensitivity to
determine subtle associations between pathways and differential methylation [@Dong2019-np;@Ren2019-mr].
The ebGSEA tool uses an empirical Bayes modeling approach followed by a one-tailed nonparametric test for gene set
enrichment [@Dong2019-np].
The `methylGSA` package implements two different approaches.
Firstly, the `methylRRA` function which uses robust rank aggregation for detecting differentially methylated genes taking
into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside each gene set
[@Ren2019-mr].
These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
direction of methylation change, it limits the utility for downstream interpretation of the importance of methylation
changes on genomic regulation.
Moreover, as pathways tend to be either upregulated or downregulated in omics assays, combining the two tends to dilute
their signal and reduce sensitivity [@Hong2014-ab].
Neither GSAmeth nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
enrichment score is a useful surrogate for effect size.
Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
biases towards large gene sets that have non-specific biological functions.

Here we aim to develop and evaluate methods for two-tailed FCS of Infinium methylation array data that addresses these
limitations.
Due to the fact that genes have a variable number of probes, there are many concevable ways that FCS could be applied
to Infinium methylation array data.
The first set of methods use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
The differential methylation of probes beloning to each gene are scored and these gene scores
undergo an enrichment test.
The second set of methods first aggregate probe methylation values for each gene before differential
methylation analysis with limma.
Then, differential gene methylation results are used for a downstream enrichment test.
With the above methods there are several methodological choices, such as the type of aggregation (eg: mean, median), and the type of
downstream test (eg: parametric or nonparametric).
To determine the best approach to applying FCS to Infinium methylation data, in this study we use
simulated data and compare results to existing ORA methods.
We then examine the sensitivity of selected methods on real cancer data, investigate the association with gene expression and end
with additional examples related to human aging and assisted reproductive technologies.

@@

Functional enrichment analysis is a process of data summarisation from genes to gene sets (pathways).
This is made more complicated for Infinium methylation array data due to the presence of multiple probes per gene,
meaning the data needs to be summarised from probes to genes and then to gene sets.
Here, we outline eight potential approaches for FCS of Infinium methylation data:

1. Limma Average t-test (LAT).
Differential methylation analysis is conducted at the probe level with limma, and the limma
t-statistics for each gene are summarised (mean).
The mean t-statistics are used in a downstream two-sample two-way t-test of gene set enrichment.

@@

4. Limma Average Mitch (LAM).
Similar to LAW, however instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
two-sample ANOVA-on-ranks test (Kruskal-Wallis).

@@

The above were compared to a standard ORA-based approach (GSA), which involves limma on probes, selection of
statistically significant probes (FDR<0.05) for ORA with the `GSAmeth()` function of missMethyl which conducts a
modified hypergeometric test that accounts for multiple probe biases.
In the case that fewer than 250 significant probes are identified, then 250 probes with the smallest p-values
were selected.
We conducted separate tests for increased and decreased probes, and we specified the background as all probes that
passed quality control filtering.

@@

To assess these methods, we adopted a simulated data approach based on the selected modification of real methylation data.
We downloaded raw intensity EPIC IDAT files from NCBI Gene Expression Omnibus (GEO) for study GSE158422 which consists of
lung tumour and normal adjacent tissues from 37 patients [@Cho2022-yd].
For the simulations, only non-cancerous datasets were used.
Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
package.
We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
3 and 12.
One thousand random gene sets were created with sizes varying between 20 and 100 members, with member genes drawn from the
annotation set.
Random gene sets were used to avoid problems caused by the large overlap among real gene sets.
Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
with decreased methylation.
From these gene sets, half of the member genes were selected.
For those selected genes, half of the annotated probes were selected.
We adjusted the M-values in the case group by a specified amount, which we call the "delta", which we varied
between 0.1 and 0.5.
Following the incorporation of methylation changes to selected probes in the case group, the data underwent
limma differential methylation analysis, followed by enrichment analysis using `GSAmeth` approach, or one of
the eight FCS methods outlined above.
Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
At each setting for group size, delta and gene set size, 100 replications were conducted each with a set seed
to ensure reproducibility.
Simulations were conducted on a SLURM cluster [@Yoo2003-mc].
Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.

@@

The full GSE158422 dataset including normal and cancer samples of all 37 patients underwent differential analysis with limma
correcting for patient-of-origin effects.
Downstream pathway enrichment analysis was conducted using the GSAmeth and LAM methods
with Reactome pathways downloaded from MSigDB website (version 2023.1)[@Milacic2024-js].
LAM relies on probe-gene associations from the "UCSC_RefGene_Name" column of the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
annotation.
As gene annotations are several years old, we used HGNChelper to update them (database current as of 21-Dec-2023) [@Oh2022-ts]
(3253 gene symbols were updated).
Euler diagrams were created with the `eulerr` R package.
To test sensitivity, a random subset of n patients was selected followed by pathway enrichment analysis with the
respective method, with n varied between 2 and 30.
Significantly enriched pathways were defined as those with FDR values < 0.05.
The subset significant results were compared to the full group (n=37) results to test the sensitivity of these enrichment
methods, and whether findings at a smaller sample size were consistent with the full group.
This process was repeated 50 times at each sample size.

@@

In order to demonstrate that methylation patterns across different array platforms can be compared using the LAM method,
we examined methylation differences in infants associated with assisted reproduction as described by two
independent studies.
The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO Accession GSE79257
[@Estill2016-jh].
The Novakovic study in was published 2019, and used the EPIC methylation array and the data is available from GEO under accession
GSE131433[@Novakovic2019-jt].
Various assisted reproductive technology conception groups are described in these studies, but we examined
just fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalistion was conducted with the SWAN method
[@Maksimovic2012-hy], and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes
located on X or Y chromosomes.
M values were computed and underwent differential methylation analysis using limma [@Ritchie2015-zn].
The Estill study included 43 naturally conceived infants and 38 with fresh embryo transfer.
The Novakovic study included 58 naturally conceived infants and 75 with fresh embryo transfer.
Limma differential analysis was conducted accounting for sex, then methylation tables underwent LAM analysis with Reactome gene sets.
Probe gene associations for HM450K array were established using the "IlluminaHumanMethylation450kanno.ilmn12.hg19" annotation set
and the gene symbols were updated with HGNChelper.
As the Novakovic study involved EPIC array data, gene-probe associations were obtained from the
"IlluminaHumanMethylationEPICanno.ilm10b4.hg19" annotation set, and gene symbols were updated with
HGNChelper.

@@

As the simulation results indicated LAM had better recall with larger gene sets, we were curious about whether there were differences
in the sizes of gene sets found by LAM and GSA.
GSA significant pathways had a median size of 135, while for LAM the median was 54.5, indicating pathways identified as significant
with LAM were collectively smaller than GSA (**Figure 3B**).
This may seem counterintuitive, but may be due to LAM's better sensitivity in dectecting pathway enrichment when effect sizes are
lower.

@@

Associations of epigenetic marks with gene expression are of great interest for understanding disease processes.
Matching tumour-normal RNA-seq datasets were analysed and we conducted enrichment analysis of gene expression together with
DNA methylation.
From 18,704 detected genes, There were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated
in the tumour group respectively.
At the pathway level there were 304 and 135 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
As the context of gene methylation is important in influencing gene expression, promoter and gene body methylation
were considered separately.
In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were were
18 and 75 pathways with higher and lower methylation respectively (FDR<0.05).
Using LAM at promoters, there were 190 and 20 pathways identified with higher and lower methylation respectively.
LAM at gene bodies identified 251 and 31 pathways with higher and lower methylation respectively (FDR<0.05).
The relative overlap between significant promoter and gene body pathways was larger for LAM pathway sets (22%) as compared to GSA
(3%) (**Figure 5A,B**).
There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).

@@

We sought to explore the utility of LAM for exploring large EWAS datasets.
One of the best studied associations with DNA methylation is with chronological aging, although
pathway level differential methylation is not well defined.
Using limma summary statistics from independent discovery and replication studies using the EPIC array platform described
by McCartney et al (2020) [@McCartney2020-ob], we applied the LAM method to explore the pathway level
differential methylation.
In the discovery cohort we observed 304 differential pathways (FDR<0.05), while in the replication group we observed 107.
There was a high degree of agreement, with 43 pathways significant in both groups.
This is not surprising, as the gene-level rank differential methylation scores show a strong positive association between discovery and
replication studies (**Figure 6A**).
Using the rank-MANOVA test of the mitch package, we identified 390 pathways with altered methylation.
Visualised as a scatterplot, the overall pattern is concordant, although a cluster of points in the lower right of the chart
indicates 268 pathways with higher methylation in the discovery sample and lower in methylation in the replication
sample (**Figure 6B**).
When prioritising the results by S distance, a surrogate measure of effect size, 29 of the top 30 pathways were concordant in their
direction of regulation (**Figure 6C**).
Some of the observed methylation changes make sense with reports in the literature.
For example the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes
beloning to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).

@@

Another application for this method is the joint pathway enrichment analysis of studies conducted with different
array systems.
Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of
studies since 2009 [@Katari2009-ya].
Two of the highest powered studies with publicly available data are Estill 2016 which used the 450K array and Novakovic 2019 study
which used the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
Joint enrichment analysis can uncover pathways that are common or different between these studies and illuminate
the biological differences between conception groups.
For this example we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation (IVF) with fresh
embryo transfer.
The Estill dataset on the 450k platform had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited
higher or lower methylation in the IVF group respectively.
The Novakovic dataset on EPIC array had 793,844 probes that met the filtering criteria (n=133).
Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibited higher or lower methylation in the IVF group
respectively.
After summarisation to the gene level, the Estill dataset on 450k array describes 19,240 genes, while the Novakovic dataset on EPIC
array describes 22,588 genes.
There were 19,234 genes common to both platforms.
The rank-rank plot of differential gene methylation scores indicates a high degree of similarity overall, and interestingly shows a
trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies
(*r*=0.35,*p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array
systems (**Figure 7B**).
A heatmap shows a high degree of agreement of pathways with higher methylation in the IVF groups,
while pathways with overall lower methylation in the IVF group showed some variability in enrichment scores
between studies (**Figure 7C**).
We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway
"Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR
value of 0.04 (**Figure 7D**).
Top ranked genes from this pathway include *ADRA2C*, *ADRB2*, *ADRA2B* and *ADRA1B*.
A loss of methylation of these genes could contribute to reported elevated blood pressure in people conceived by IVF [@Meister2018-cp].
Taken together, this analysis shows that LAM method can be used to compare enrichment of independent studies
conducted with 450k and EPIC arrays.

![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
Red indicates higher methylation and blue shows lower methylation.
(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
member genes in both Estill and Novakovic studies.](img/fig7.svg)

@@

To examine the potential for LAM to illuminate differential pathway methylation associations
with diseases, we obtained and analysed blood based EWAS summary statistics from a recent
publication [@Hillary2023-mu].
This study examined 14 prevalent disease states and incidence of 19 disease states in a group
of 18,413 participants, using the EPIC array.
Delta-beta values were used for the purpose of scoring probe differential methylation, and these
values underwent LAM analysis with Reactome gene sets.

>>>>

<strong>
Mark Ziemann<sup>1,2*</sup>,
Mandhri Abeysooriya<sup>2,3</sup>,
Anusuiya Bora<sup>1,2</sup>,
Séverine Lamon<sup>3</sup>,
Mary Sravya Kasu<sup>1</sup>,
Mitch W. Norris<sup>1</sup>,
Yen Ting Wong<sup>4</sup>,
Jeffrey M. Craig<sup>4,5</sup>
</strong>

@@

Infinium methylation bead arrays remain one of the most popular platforms for studying genomic DNA methylation,
but tools for downstream pathways analysis have their limitations.
Functional class scoring (FCS) is a group of pathway enrichment techniques that involve the ranking of genes
and evaluation of their collective regulation in biological systems, but the implementations described for
Infinium methylation data do not retain direction information which is important for mechanistic understanding
of genomic regulation.
Here, we evaluate several candidate FCS methods that retain directional information.
According to simulation results, the best performing method involves the mean aggregation of probe limma
t-statistics by gene followed by a rank-ANOVA enrichment test using the mitch package.
This method, which we call “LAM”, outperformed an existing over-representation analysis method in simulations,
and showed higher sensitivity and robustness in an analysis of real lung tumour-normal paired datasets.
Using matched RNA-seq data we examine the relationship of methylation differences at promoters and gene bodies
with RNA expression at the level of pathways in lung cancer.
To demonstrate the utility of our approach, we apply it to three other contexts where public data were available.
Firstly we examine differential pathway methylation associated with chronological age.
Secondly we investigate pathway methylation differences in infants conceived with in vitro fertilisation.
Lastly, we analyse differential pathway methylation in 19 disease states, identifying hundreds of novel associations.
These results show LAM is a powerful method for the detection of differential pathway methylation as compared to
existing methods.
A reproducible vignette is provided to illustrate how to implement this method.

@@

DNA methylation is the most widely studied epigenetic mark, for its role in
development and disease [@Robertson2005-tu].
Hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
methylation patterns in disease and other contexts [@Li2019-og].
Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
as compared to high throughput methylation sequencing [@Wei2021-qj].
Infinium arrays typically include multiple probes per gene in different locations including promoters,
CpG islands, gene bodies and enhancers [@Pidsley2016-ix], which complicates downstream functional enrichment analysis,
otherwise known as pathway analysis or gene set enrichment analysis.
Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
class scoring (FCS) [@Khatri2012-qy].
A third type of enrichment analysis called pathway topology improves upon other methods with more sophisticated
modelling of gene network activations, but these are yet to be adopted widely [@Xie2021-py].
In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
to a background list of all genes measured in the assay.
The test seeks to identify sets of genes (eg: ontologies, pathways) that are over-represented in the
gene list of interest relative to the background [@Tavazoie1999-ng].
FCS takes a different approach by ranking all detected genes by a differential
regulation score (eg: fold change, confident effect size, t-statistic) followed by a test to assess whether each
set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of infinium methylation array
data as it addresses issues related to probes belonging to more than one gene and the fact that one gene can have multiple
probes [@Maksimovic2021-ur].
As gsameth is an ORA method, results strongly depend on the significance threshold used [@Ren2019-mr], and also
on the proportion of probes that meet this threshold.
FCS methods `ebGSEA` and `methylGSA` have been developed for EWAS data and are suggested to have better sensitivity to
determine subtle associations between pathways and differential methylation [@Dong2019-np;@Ren2019-mr].
The ebGSEA tool uses an empirical Bayes modeling approach followed by a one-tailed nonparametric test for gene set
enrichment [@Dong2019-np].
The `methylGSA` package implements two different approaches.
Firstly, the `methylRRA` function, which uses robust rank aggregation for detecting differentially methylated genes taking
into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside each gene set
[@Ren2019-mr].
These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
direction of methylation change, it limits the utility for downstream interpretation of the importance of methylation
changes on genomic regulation.
Moreover, as pathways tend to be either upregulated or downregulated in omics assays, combining the two tends to dilute
their signal and reduce sensitivity [@Hong2014-ab].
Neither GSAmeth nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
enrichment score is a useful surrogate for effect size.
Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
biases towards large gene sets that have non-specific biological functions.

Here we aim to develop and evaluate methods for two-tailed FCS of Infinium methylation array data that addresses these
limitations.
Since genes have a variable number of probes, there are many concevable ways that FCS could be applied
to Infinium methylation array data.
The first set of methods (type I) use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
The differential methylation of probes beloning to each gene are scored and these gene scores
undergo an enrichment test.
The second set of methods (type II) first aggregate probe methylation values for each gene before differential
methylation analysis with limma.
Then, differential gene methylation results are used for a downstream enrichment test.
With the above methods there are several methodological choices, such as the type of aggregation, and the type of
downstream test.
To determine the best approach to applying FCS to Infinium methylation data, in this study, we use
simulated data and compare results to existing ORA methods.
We then examine the sensitivity of selected methods on real cancer data, investigate the association with gene expression
and end with additional examples related to human ageing, assisted reproductive technologies and a large-scale EWAS of
19 diseases.

@@

Functional enrichment analysis is a process of data summarisation from genes to gene sets (pathways).
This is made more complicated for Infinium methylation array data due to the presence of multiple probes per gene,
meaning the data needs to be summarised from probes to genes and then to gene sets.
Here, we outline eight potential approaches for FCS of Infinium methylation data, where methods 1-5 are type I
and 6-8 are type II:

1. Limma Average t-test (LAT).
Differential methylation analysis is conducted at the probe level with limma, and the limma
t-statistics for each gene are summarised (arithmetic mean).
The mean t-statistics are used in a downstream two-sample two-way t-test of gene set enrichment.

@@

4. Limma Average Mitch (LAM).
Similar to LAW, however instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
two-sample ANOVA-on-ranks test as described previously[@Cox2012-zu].

@@

The above were compared to a standard ORA-based approach (GSA), which involves limma on probes, selection of
statistically significant probes (FDR<0.05) for ORA with the `GSAmeth()` function of missMethyl which conducts a
modified hypergeometric test that accounts for multiple probe biases.
In the case that fewer than 250 significant probes were identified, the 250 probes with the smallest p-values
were selected.
We conducted separate tests for increased and decreased probes, and we specified the background as all probes that
passed quality control filtering.

@@

To assess these methods, we adopted a simulated data approach based on the selected modification of real methylation data.
We downloaded raw intensity EPIC IDAT files from NCBI Gene Expression Omnibus (GEO) for study GSE158422, which consists of
lung tumour and normal adjacent tissues from 37 patients [@Cho2022-yd].
For the simulations, only non-cancerous datasets were used.
Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
package.
We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
3 and 12.
One thousand random gene sets were created with sizes varying between 20 and 100 members, with member genes drawn from the
annotation set.
Random gene sets were used to avoid problems caused by the large overlap among real gene sets.
Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
with decreased methylation.
From these gene sets, half of the member genes were selected.
For those selected genes, half of the annotated probes were selected.
We adjusted the M-values in the case group by a specified amount, which we call the "delta", which we varied
between 0.1 and 0.5.
Following the incorporation of methylation changes to selected probes in the case group, the data underwent
limma differential methylation analysis, followed by enrichment analysis using the `GSAmeth` approach, or one of
the eight FCS methods outlined above.
Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
At each setting for group size, delta and gene set size, 100 replications were conducted, each with a set seed
to ensure reproducibility.
Simulations were conducted on a SLURM cluster [@Yoo2003-mc].
Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.

@@

The full GSE158422 dataset including normal and cancer samples of all 37 patients underwent differential analysis with limma
correcting for patient-of-origin effects.
Downstream pathway enrichment analysis was conducted using the GSAmeth and LAM methods
with Reactome pathways downloaded from MSigDB website (version 2023.1)[@Milacic2024-js].
LAM relies on probe-gene associations from the "UCSC_RefGene_Name" column of the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
annotation.
As gene annotations are several years old, we used HGNChelper to update them (database current as of 21-Dec-2023) [@Oh2022-ts]
(3,253 gene symbols were updated).
Euler diagrams were created with the `eulerr` R package.
To test sensitivity, a random subset of n patients was selected followed by pathway enrichment analysis with the
respective method, with n varying between 2 and 30.
Significantly enriched pathways were defined as those with FDR values < 0.05.
The subset significant results were compared to the full group (n=37) results to test the sensitivity of these enrichment
methods, and whether findings at a smaller sample size were consistent with the full group.
This process was repeated 50 times for each sample size.

@@

In order to demonstrate that methylation patterns across different array platforms can be compared using the LAM method,
we examined methylation differences in infants associated with assisted reproduction as described by two
independent studies.
The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO Accession GSE79257
[@Estill2016-jh].
The Novakovic study in was published 2019, and used the EPIC methylation array and the data is available from GEO under accession
GSE131433[@Novakovic2019-jt].
Various assisted reproductive technology conception groups are described in these studies, but we only examined
fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalistion was conducted with the SWAN method
[@Maksimovic2012-hy], and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes
located on X or Y chromosomes.
M values were computed and underwent differential methylation analysis using limma [@Ritchie2015-zn].
The Estill study included 43 naturally conceived infants and 38 with fresh embryo transfer.
The Novakovic study included 58 naturally conceived infants and 75 with fresh embryo transfer.
Limma differential analysis was conducted accounting for sex, then methylation tables underwent LAM analysis with Reactome gene sets.
Probe gene associations for HM450K array were established using the "IlluminaHumanMethylation450kanno.ilmn12.hg19" annotation set
and the gene symbols were updated with HGNChelper.
As the Novakovic study involved EPIC array data, gene-probe associations were obtained from the
"IlluminaHumanMethylationEPICanno.ilm10b4.hg19" annotation set, and gene symbols were updated with
HGNChelper.

@@

As the simulation results indicated LAM had better recall with larger gene sets, we were curious about whether there were differences
in the sizes of gene sets found by LAM and GSA.
GSA significant pathways had a median size of 135, while for LAM the median was 54.5, indicating pathways identified as significant
with LAM were collectively smaller than GSA (**Figure 3B**).
This may seem counterintuitive but may be due to LAM's better sensitivity in dectecting pathway enrichment when effect sizes are
lower.

@@

Associations of epigenetic marks with gene expression are of great interest for understanding disease processes.
Matching tumour-normal RNA-seq datasets were analysed and we conducted enrichment analysis of gene expression together with
DNA methylation.
From 18,704 detected genes, there were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated
in the tumour group respectively.
At the pathway level there were 304 and 135 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
As the context of gene methylation is important in influencing gene expression, promoter and gene body methylation
were considered separately.
In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were were
18 and 75 pathways with higher and lower methylation respectively (FDR<0.05).
Using LAM at promoters, there were 190 and 20 pathways identified with higher and lower methylation respectively.
LAM at gene bodies identified 251 and 31 pathways with higher and lower methylation respectively (FDR<0.05).
The relative overlap between significant promoter and gene body pathways was larger for LAM pathway sets (22%) as compared to GSA
(3%) (**Figure 5A,B**).
There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).

@@

We sought to explore the utility of LAM for exploring large EWAS datasets.
One of the best studied associations with DNA methylation is with chronological aging, although
pathway level differential methylation is not well defined.
Using limma summary statistics from independent discovery and replication studies using the EPIC array platform described
by McCartney et al (2020) [@McCartney2020-ob], we applied the LAM method to explore the pathway level
differential methylation.
In the discovery cohort we observed 304 differential pathways (FDR<0.05), while in the replication group we observed 107.
There was a high degree of agreement, with 43 pathways significant in both groups.
This is not surprising, as the gene-level rank differential methylation scores show a strong positive association between discovery and
replication studies (**Figure 6A**).
Using the rank-MANOVA test of the mitch package, we identified 390 pathways with altered methylation.
Visualised as a scatterplot, the overall pattern is concordant, although a cluster of points in the lower right of the chart
indicates 268 pathways with higher methylation in the discovery sample and lower in methylation in the replication
sample (**Figure 6B**).
When prioritising the results by S distance, a surrogate measure of effect size, 29 of the top 30 pathways were concordant in their
direction of regulation (**Figure 6C**).
Some of the observed methylation changes make sense with reports in the literature.
For example, the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes
beloning to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).

@@

Another application for this method is the joint pathway enrichment analysis of studies conducted with different
array systems.
Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of
studies since 2009 [@Katari2009-ya].
Two of the highest powered studies with publicly available data are Estill 2016 which used the HM450K array and Novakovic 2019 study
which used the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
Joint enrichment analysis can uncover pathways that are common or different between these studies and illuminate
the biological differences between conception groups.
For this example we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation (IVF) with fresh
embryo transfer.
The Estill dataset on the HM450K platform had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited
higher or lower methylation in the IVF group respectively.
The Novakovic dataset on EPIC array had 793,844 probes that met the filtering criteria (n=133).
Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibited higher or lower methylation in the IVF group
respectively.
After summarisation to the gene level, the Estill dataset on HM450K array describes 19,240 genes, while the Novakovic dataset on EPIC
array describes 22,588 genes.
There were 19,234 genes common to both platforms.
The rank-rank plot of differential gene methylation scores indicates a high degree of similarity overall, and interestingly shows a
trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies
(*r*=0.35, *p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array
systems (**Figure 7B**).
A heatmap shows a high degree of agreement of pathways with higher methylation in the IVF groups,
while pathways with overall lower methylation in the IVF group showed some variability in enrichment scores
between studies (**Figure 7C**).
We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway
"Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR
value of 0.04 (**Figure 7D**).
Top ranked genes from this pathway include *ADRA2C*, *ADRB2*, *ADRA2B* and *ADRA1B*.
A loss of methylation of these genes could contribute to reported elevated blood pressure in people conceived by IVF [@Meister2018-cp].
Taken together, this analysis shows that LAM method can be used to compare enrichment of independent studies
conducted with HM450K and EPIC arrays.

![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (HM450K) and Novakovic (EPIC) studies.
(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
Red indicates higher methylation and blue shows lower methylation.
(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
member genes in both Estill and Novakovic studies.](img/fig7.svg)

@@

To examine the potential for LAM to reveal differential pathway methylation associations
with diseases, we obtained and analysed blood based EWAS summary statistics from a recent
publication [@Hillary2023-mu].
This study examined 14 prevalent disease states and incidence of 19 disease states in a group
of 18,413 participants, using the EPIC array.
Delta-beta values were used for the purpose of scoring probe differential methylation, and these
values underwent LAM analysis with Reactome gene sets.
